Self-identification of gender and sexual orientation in the US General Social Survey:

### 2022-03-27 paper_3.Rmd
near end of work mar 26

---
title: "Self-identification of gender and sexual orientation in the US General Social Survey:"
subtitle: "A review of methodological inconsistencies and anomalous data"
abstract: >
  The US General Social Survey, administered by the National Opinion Research Center at the University of Chicago, was recently emended to include options for respondents to specify several aspects of their sexual orientation and gender identity, including how their gender has changed relative to their birth sex. However, upon exploration of the data, it became clear that several major inconsistencies and anomalies existed in both the methodology by which the data were collected, as well as in the data themselves. This paper highlights these issues, and offers as an appendix a supplementary survey that would help to rectify these anomalies in future General Social Surveys.
author: "Oliver Daniel & Tiago Martins"
date: "due March 20, 2022[^1]"
geometry: "margin=1.5in"
knit: (function(input, ...) {
    rmarkdown::render(
    input,
    output_file = "out/paper_3.pdf",
    envir = globalenv(),
    ...
    )
  })
bibliography: in/references.bib
link-citations: true
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = "80%", out.align="center", dir)
options(knitr.graphics.error = FALSE)
```

```{r libraries, include=FALSE}
if (!require(tidyverse)) {install.packages('tidyverse')}
if (!require(gssr)) {remotes::install_github('https://github.com/kjhealy/gssr')}

library(tidyverse)
library(gssr)
```

```{r data-setup}
data(gss_all)
gss_clean <- gss_all |>
  mutate(
    id = paste0(year, '_', id),
    sexnow = coalesce(sexnow, sexnow1),
    sexbirth = coalesce(sexbirth, sexbirth1),
    sexsex = sexsex,
    sexsex5 = sexsex5,
    ornt_sex_label = paste(
      case_when(
        sexornt == 1 ~ "Gay",
        sexornt == 2 ~ "Bisexual",
        sexornt == 3 ~ "Straight"
      ),
      case_when(
        sexnow == 1 ~ "man",
        sexnow == 2 ~ "woman",
        sexnow == 3 ~ "trans person",
        sexnow == 4 ~ "person of another gender",
      )
    ),
    .keep = "used"
  )
```

@@

## Introduction
TODO
## Data
### Dataset
This paper focuses on the United States General Social Survey (US-GSS, or GSS in the the following), a survey regularly administered by the National Opinion Research Center (NORC) at the University of Chicago [@norc-gss]. For reasons elaborated below, although the cumulative data file contains `r nrow(gss_clean) |> format(big.mark=',')` observations across `r gss_clean$year |> n_distinct()` years of study, we will be focusing primarily on only respondents from the last two surveys, in 2018 and 2021. This significantly limits certain aspects of our investigation to `r gss_clean |> filter(year >= 2018) |> nrow() |> format(big.mark =',')` entries; however, the issues we highlight are endemic to the format and methodology of the survey itself, and thus do not require large magnitudes to afford predictive power.

@@

- `id`: A serial ID. Not used directly, but it was discovered that the serial IDs for respondents rolledback to 1 every survey year. For reproducibility, it was important to ensure that there were no ID collisions. So, the serial ID is now also prefixed with the year.
- `sex_ornt_label`: A compact, human-readable description of a person's gender and sexual orientation, as might be used in everyday conversation. e.g., "Straight woman", "Gay man", "Bisexual trans person".[^4]

### Methods
All data analysis[^2] was performed in the R statistical programming language [@R], using RStudio [@RStudio] for ease of exploration, development, and production of this paper via built-in `knitr` [@knitr] functionality. GSS data are fetched by the `gssr` package [@gssr], which automatically packages the columns of the data into a ready-to-use dataframe. As usual, the `tidyverse` [@tidyverse] collection provides an indispensable array of resources, including but not limited to `ggplot2` [@ggplot2] for visualization.

## Results
TODO
## Discussion
TODO
## Survey
TODO

@@

<!-- Footnotes -->
[^1]: Extended.
[^2]: Code available at https://github.com/oliver-daniel/inf313_paper_3.
[^3]: There is an important distinction to be made between sexual attraction and romantic attraction, not to mention sexual attraction valency (i.e., the asexuality spectrum). However, for the purposes of this study and supplementary survey, only the prior is considered.
[^4]: Note that "gay" is used to refer to any kind of exclusively-homosexual attraction, as identified by the respondent. Historically, "gay" often referred exclusively to male homosexual attraction, with "lesbian" being more common for females; conversely, "gay" is currently entering common parlance for *any* same-gender attraction, even among those who also experience different-gender attraction. The usage of terms such as these to refer to different intersections of gender and sexuality over time is fascinating, and worthy of its own study; however, it is immaterial to this study, and so "gay" was chosen as a cover-all term.

>>>>

---
title: "Self-identification of gender and sexual orientation in the US General Social Survey:"
subtitle: "A review of methodological inconsistencies and anomalous data"
abstract: >
  The US General Social Survey, administered by the National Opinion Research Center at the University of Chicago, was recently emended to include options for respondents to specify several aspects of their sexual orientation and gender identity, including how their gender has changed relative to their birth sex. However, upon exploration of the data, it became clear that several major inconsistencies and anomalies existed in both the methodology by which the data were collected, as well as in the data themselves. This paper highlights these issues, and offers as an appendix a supplementary survey that would help to rectify these anomalies in future General Social Surveys.
author: "Oliver Daniel & Tiago Martins"
date: "due March 20, 2022[^1]"
geometry: "margin=1.5in"
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: false
    fig_crop: false
    extra_dependencies: ["flafter"]
knit: (function(input, ...) {
    rmarkdown::render(
    input,
    output_file = "out/paper_3.pdf",
    envir = globalenv(),
    ...
    )
  })
bibliography: in/references.bib
link-citations: true
---

```{r preamble}
#### Preamble ####
# Purpose: Fetch, clean, and present data from the NORC United States General Social Survey, knitted into a research paper
# Authors: Oliver Daniel and Tiago Martins
# Date: 2022-03-20
# Contact: via Quercus or email
# License: WTFPL
# Prerequisites:
# - None known; running this file should auto-install
# dependencies.

#TODO: move setup to another script?
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, out.width = "80%", out.align="center", dir)
options(knitr.graphics.error = FALSE)
```

```{r libraries, include=FALSE}
if (!require(tidyverse)) {install.packages('tidyverse')}
if (!require(gssr)) {remotes::install_github('https://github.com/kjhealy/gssr')}
if (!require(kableExtra)) {install.packages('kableExtra')}
if (!require(patchwork)) {install.packages('patchwork')}

library(tidyverse)
library(kableExtra)
library(patchwork)
library(gssr)
```

```{r data-setup}
data(gss_all)
gss_clean <- gss_all |>
  mutate(
    id = paste0(year, '_', id),
    sex = sex,
    sexnow = coalesce(sexnow, sexnow1),
    sexbirth = coalesce(sexbirth, sexbirth1),
    sexsex = sexsex,
    sexsex5 = sexsex5,
    ptnrornt = ptnrornt,
    ornt_sex_label = if_else(
      is.na(sexornt) | is.na(sexnow),
      as.character(NA),
      paste(
      case_when(
        sexornt == 1 ~ "Gay",
        sexornt == 2 ~ "Bisexual",
        sexornt == 3 ~ "Straight"
      ),
      case_when(
        sexnow == 1 ~ "men",
        sexnow == 2 ~ "women",
        sexnow == 3 ~ "trans people",
        sexnow == 4 ~ "people of other genders",
      )
    )),
    .keep = "used"
  ) |>
  select(-one_of(c('sexnow1', 'sexbirth1')))
```

@@

```{r colours}
colour <- list(
  male = '#0273f5',
  intersex = '#9B02ED',
  female = '#fb0036',
  trans = '#F495D1',
  other = '#098000',
  gay = '#A50314',
  bisexual = '#B759D9',
  straight = '#0375A6',
  none = '#464646'
)

```

## Introduction
TODO

## Data
### Dataset
This paper focuses on the United States General Social Survey (US-GSS, or GSS in the the following), a survey regularly administered by the National Opinion Research Center (NORC) at the University of Chicago [@norc-gss]. For reasons elaborated below, although the cumulative data file contains `r nrow(gss_clean) |> format(big.mark=',')` observations across `r gss_clean$year |> n_distinct()` years of study, we will be focusing primarily on only respondents from the last two surveys, in 2018 and 2021. This significantly limits certain aspects of our investigation to `r gss_clean |> filter(year >= 2018) |> nrow() |> format(big.mark =',')` entries; however, the issues we highlight are endemic to the format and methodology of the survey itself, and thus do not require large magnitudes to afford predictive power.

@@

- `id`: A serial ID. Not used directly, but it was discovered that the serial IDs for respondents rolledback to 1 every survey year. For reproducibility, it was important to ensure that there were no ID collisions. So, the serial ID is now also prefixed with the year.
- `sex_ornt_label`: A compact, human-readable description of a person's gender and sexual orientation, as might be used in everyday conversation. e.g., "Straight woman", "Gay man", "Bisexual trans person".[^5]

### Methods
All data analysis[^2] was performed in the R statistical programming language [@R], using RStudio [@RStudio] for ease of exploration, development, and production of this paper via built-in `knitr` [@knitr] functionality. GSS data are fetched by the `gssr` package [@gssr], which automatically packages the columns of the data into a ready-to-use dataframe. As usual, the `tidyverse` [@tidyverse] collection provides an indispensable array of resources, including but not limited to `ggplot2` [@ggplot2] for visualization. For additional aesthetics, `kableExtra` [@kableExtra], which[^6] styles tables, and `patchwork` [@patchwork], which composes multiple `ggplot` plots together, were also used.

## Results

```{r fig-demographics, fig.cap="\\label{fig:demographics} In both 2018 and 2021, respondent demographics are dominated by straight men and women. A significantly larger contingent of bisexual women responded in 2021."}
subsetWith('year', 'sexnow', 'sexornt', sel = T) |>
  ggplot(aes(x = as_factor(sexnow), y = after_stat(count), fill = as_factor(sexornt))) +
  geom_bar() +
  facet_wrap(~year) +
  theme(
    axis.text.x = element_text(angle=20, vjust=.95, hjust=1),
    axis.title.y = element_blank(),
    legend.position = "bottom",
    legend.direction = "vertical"
  ) +
  labs(
    title = "Gender and sexuality identification demographics of GSS respondents across 2018 and 2021 surveys" |> str_wrap(60),
    x = "Gender identity of respondent",
    fill = "Sexual orientation of respondent"
  ) +
  scale_fill_manual(values=c(colour$gay, colour$bisexual, colour$straight))
```
```{r fig-demographics-trans-nb, fig.cap="\\label{fig:demographics-trans-nb} The overall turnout of transgender and 'other' (e.g., non-binary) respondents is low, but exhibits a wider diversity in sexuality."}
subsetWith('year', 'sexnow', 'sexornt', sel = T) |>
  filter(sexnow > 2) |>
  ggplot(aes(x = as_factor(sexnow), y = after_stat(count), fill = as_factor(sexornt))) +
  geom_bar() +
  facet_wrap(~year) +
  theme(
    axis.text.x = element_text(angle=20, vjust=.95, hjust=1),
    axis.title.y = element_blank(),
    legend.position = "bottom"
  ) +
  labs(
    title = "Demographics of transgender and 'other' GSS respondents across 2018 and 2021 surveys" |> str_wrap(60),
    x = "Gender identity of respondent",
    fill = "Sexual orientation of respondent"
  ) +
  scale_fill_manual(values=c(colour$gay, colour$bisexual, colour$straight))
```

```{r table-birth-now}

counts <- 
  subsetWith('sexbirth', 'sexnow', sel=T) |> 
  table()

rownames(counts) <- c('Assigned male', 'Assigned female', 'Intersex')

knitr::kable(
    counts,
    col.names = c('Male', 'Female', 'Transgender', 'Other'),
    caption = "As expected, cisgendered respondents vastly outnumber any other configuration; in this dataset, the entire intersex population of the United States is represented by a single respondent."
  )
```
More than 22% of self-reported cisgender women are recorded as male, and 30% of self-reported cisgender men as female.
```{r table-sex-coding}
birth_codes <- c('M', 'F', 'I')
gender_codes <- c('M', 'F', 'T', 'X')

counts <- subsetWith('sexbirth', 'sexnow') |>
  mutate(
    code = paste0(birth_codes[sexbirth], '/', gender_codes[sexnow]),
    ) |>
  select(code, sex) |>
  table()

colnames(counts) <- c('Recorded as male', 'Recorded as female')

  knitr::kable(
    counts,
    caption="This table exhibits the first major inconsistency in the data."
    )
```

```{r fig-latest-sexual-partners, fig.cap="\\label{fig:latest-sexual-partners}Surprisingly, the distributions of this variable seem to be more accurately reflective than those in Figure \\ref{fig:sexual-partners-over-time}, in terms of what sorts of sexual partners one would expect a respondent to have had."}
makePlot <- function(y_var) {
  subsetWith('ornt_sex_label', all_of(y_var), sel=T) |>
    ggplot(aes(y = reorder(ornt_sex_label, ornt_sex_label, length), fill = as_factor(.data[[all_of(y_var)]]))) +
    geom_bar(position = position_fill(reverse=TRUE))
}

makePlot('ptnrornt') + theme(legend.position = "bottom", legend.direction = "vertical", axis.title.y = element_blank(), axis.title.x = element_blank(), legend.title = element_blank(), legend.justification = "left", axis.ticks.y = element_blank()) + plot_annotation(
  title = '"In terms of their orientation, my latest sexual partner was..."',
  subtitle = "by personal identity, as a proportion of respondents"
) + scale_fill_manual(values=c(colour$gay, colour$bisexual, colour$straight, colour$none))
```

```{r fig-sexual-partners-over-time, fig.cap="\\label{fig:sexual-partners-over-time}Although preference in sexual partners appears relatively stable over both 1-year and 5-year periods, an anomalously large contingent of heterosexual-identifying men and women have had exclusively same-sex relationships over these periods."}
(makePlot('sexsex') + labs(title = "over the past 12 months")) +
(makePlot('sexsex5') + labs(title = "over the past 5 years") + theme(axis.text.y = element_blank())) +
  plot_layout(guides = "collect")  +
  plot_annotation(title = '"In terms of their sex, my sexual partners have been..."',
  subtitle = "by personal identity, as a proportion of respondents") &
  theme(legend.position = "bottom", axis.title.x = element_blank(), axis.title.y = element_blank(), legend.title = element_blank(), axis.ticks.y = element_blank()) &
  scale_fill_manual(values=c(colour$male, colour$bisexual, colour$female))
```

## Discussion
TODO

## Survey
TODO

@@

<!-- Footnotes -->
[^1]: Extended.
[^2]: Code available at https://github.com/oliver-daniel/inf313_paper_3.
[^3]: There is an important distinction to be made between sexual attraction and romantic attraction, not to mention sexual attraction valency (i.e., the asexuality spectrum). However, for the purposes of this study and supplementary survey, only the prior is considered.
[^4]: This is the only variable under investigation which includes an option for "I have never had a sexual partner". As a result, most analyses involving this variable will explicitly filter out this option, to keep a level playing field among the other sex-partner-related questions.
[^5]: Note that "gay" is used to refer to any kind of exclusively-homosexual attraction, as identified by the respondent. Historically, "gay" often referred exclusively to male homosexual attraction, with "lesbian" being more common for females; conversely, "gay" is currently entering common parlance for *any* same-gender attraction, even among those who also experience different-gender attraction. The usage of terms such as these to refer to different intersections of gender and sexuality over time is fascinating, and worthy of its own study; however, it is immaterial to this study, and so "gay" was chosen as a cover-all term.
[^6]: Attempts to, anyway. Half of the options cause Kable to dump raw \LaTeX  and destroy the formatting of the entire document. Did you know that it's literally impossible to move table captions to the bottom of the table?
